Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2016-12-20 draft.tex
.

Sweeney et al. found that diffusive homeostasis maintains broadness of firing rates across a wide range of diffusion constants but rapidly approaching zero for small values \cite[p. 6]{Sweeney_Paper}. We were able to reproduce this result, see Figure \ref{Fir_Rate_Dist_Width_vs_D}. Homeostasis reaches a point of saturation, where faster diffusion has no effect on the heterogeneity of firing rates. 
\begin{figure}
%%% freq_dist_width_vs_D.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/std_dev_vs_D_neumann.png}
\caption{Standard deviation of firing rate distribution of excitatory neurons (neumann boundary conditions).}
\label{Fir_Rate_Dist_Width_vs_D}
\end{figure}
To further quantify this dependence, we also investigated the influence of the diffusion constant onto the distribution's skewness, shown in Figure \ref{Fir_Rate_Dist_Skew_vs_D}. Compared to the standard deviation, we see a similar but not as clear trend with a drop for very small diffusion constants, even occasionally resulting in a left-skewed distribution (negative D-values).
\begin{figure}
%%% freq_dist_skew_vs_D.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/skew_vs_D_neumann.png}
\caption{Skewness of firing rate distribution of excitatory neurons (neumann boundary conditions).}
\label{Fir_Rate_Dist_Skew_vs_D}
\end{figure}
A naturally emerging question when altering the diffusion constant is how the firing rate behaves in the absolute limit of infinitely fast diffusion. In fact, simulation-wise this case is quite easy to simulate: one simply has  to feed all NO-sources into a single scalar variable of NO concentration. In particular, this will provide the same NO readout for all excitatory neurons, which means that all excitatory thresholds change at the same rate all the time. Since we initialized all excitatory neurons at the same threshold, only one global excitatory threshold variable had to be considered in this case. Figure \ref{Fir_Rate_Dist_Instant_compare} shows the distribution for this special limiting case. The standard deviation for the excitatory was $\mathrm{\sigma_{inst} = 1.41\, Hz}$ and the skewness $\mathrm{v_{inst} = 1.27}$ ($\mathrm{v_{inst,log} = -0.46}$), which makes the asymmetry slightly more pronounced than in Figure \ref{Fir_Rate_Dist_Compare}.  
\begin{figure}
%%% frequ_dist_instant_diff.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_e.png}
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_i.png}
\caption{Distribution of firing rates for instantaneous diffusion for excitatory (A,B) and inhibitory (A*,B*) population. Data was taken from 2 simulations and $\mathrm{t=1000-1500\,s}$.}
\label{Fir_Rate_Dist_Instant_compare}
\end{figure}

Summing up the results of this section, we can state that stable network activity can be achieved in the SORN with diffusive homeostasis while producing a broad distribution of firing rates within the excitatory population. We found that this ensemble of activity is maintained on a single cell level, each cell spiking with approximately poissonian statistics at a constant mean rate. The broadness of distribution could be maintained at a relatively constant level over a wide range of diffusion rates. However, we did not find a distribution as right skewed and heavy tailed as reported by Sweeney et al. and numerous experimental studies.

\subsection{Network Topology}
So far, we have only analyzed network activity, whereas in this section we present results that relate to features of synaptic topology. Since the SORN we used for all simulations only included plastic mechanisms affecting recurrent excitatory connections, these were the main subject of our investigations. Moreover, in contrast to the previous section, we did not aim for the emergence of an explicitly \emph{new} feature of synaptic topology, but rather sought to recover properties that had been found in earlier versions of the SORN.

\subsubsection{Excitatory Connection Fraction}
In the previous version of the SORN, initializing the network with zero recurrent excitatory connections led to a monotonically increasing but saturating excitatory-to-excitatory connection fraction (CF). The growth rate was tuned such that the terminal CF settled at $\mathrm{10\%}$, see \cite[p. 8]{SORN_Paper}. Omitting the distance dependence of connection probability for synaptic growth led to a slightly higher connection fraction. To compare these previous results to the diffusive case, we ran a full simulation under diffusive homeostasis, including the growth phase by setting the target concentration to an appropriate value that was determined in a previous simulation with non-diffusive homeostasis. The resulting time course of CF is depicted in Figure \ref{CF_plot}.
\begin{figure}
%%% plot_CF.py
\includegraphics[width=\textwidth]{../../plots/CF_plot.png}
\caption{Connection fractions of recurrent excitatory synapses. Simulation protocols include all four possible combinations of diffusive/non-diffusive homeostasis and distance-dependent/non-distance-dependent synaptic growth, each variant retained for the entire simulation time.}
\label{CF_plot}
\end{figure}
To our surprise, diffusive homeostasis caused the CF to overshoot its final fraction in the growth phase. Eventually however, it settled at a similar equilibrium of $\mathrm{\approx 10\%}$. To understand this kind of behaviour, we recalled the two mechanisms that directly determine changes within the connection fraction, namely synaptic growth and pruning. The average change of connection fraction is simply proportional to the difference between newly created synapses and those that were removed. Though fluctuating, the synaptic growth rate was kept at a constant level of 920 synapses/second. In consequence, any differences in Figure \ref{CF_plot} must originate in different pruning rates. As explained in section \ref{network simulation}, pruning is performed once per second by collection connections that have fallen under the given weight-threshold. Within this second, a synaptic weight can fall below this value for two reasons: either due to multiplicative normalization or additive STDP. If the weight of a neuron is already very small, multiplicative changes due to normalization have a minor effect compared to weight fluctuations caused by additive STDP, drawing us to the conclusion that differences within the statistics of STDP are the potential cause for the observed overshooting of the CF.

>>>>

Sweeney et al. found that diffusive homeostasis maintains broadness of firing rates across a wide range of diffusion constants but rapidly approaching zero for small values \cite[p. 6]{Sweeney_Paper}. We were able to reproduce this result, see Figure \ref{Fir_Rate_Dist_Width_vs_D}. Homeostasis reaches a point of saturation, where faster diffusion has no effect on the heterogeneity of firing rates. 
\begin{figure}
%%% freq_dist_width_vs_D.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/std_dev_vs_D_neumann.png}
\caption{Standard deviation of firing rate distribution of excitatory neurons (neumann boundary conditions).}
\label{Fir_Rate_Dist_Width_vs_D}
\end{figure}
To further quantify this dependence, we also investigated the influence of the diffusion constant onto the distribution's skewness, shown in Figure \ref{Fir_Rate_Dist_Skew_vs_D}. Compared to the standard deviation, we see a similar but not as clear trend with a drop for very small diffusion constants, even occasionally resulting in a left-skewed distribution (negative D-values).
\begin{figure}
%%% freq_dist_skew_vs_D.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/skew_vs_D_neumann.png}
\caption{Skewness of firing rate distribution of excitatory neurons (neumann boundary conditions).}
\label{Fir_Rate_Dist_Skew_vs_D}
\end{figure}
A naturally emerging question when altering the diffusion constant is how the firing rate behaves in the absolute limit of infinitely fast diffusion. In fact, simulation-wise this case is quite easy to simulate: one simply has  to feed all NO-sources into a single scalar variable of NO concentration. In particular, this will provide the same NO readout for all excitatory neurons, which means that all excitatory thresholds change at the same rate all the time, only shifting their initial random distribution. Figure \ref{Fir_Rate_Dist_Instant_compare} shows the distribution for this special limiting case. The standard deviation for the excitatory was $\mathrm{\sigma_{inst} = 1.41\, Hz}$ and the skewness $\mathrm{v_{inst} = 1.27}$ ($\mathrm{v_{inst,log} = -0.46}$), which makes the asymmetry slightly more pronounced than in Figure \ref{Fir_Rate_Dist_Compare}.  
\begin{figure}
%%% frequ_dist_instant_diff.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_e.png}
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_i.png}
\caption{Distribution of firing rates for instantaneous diffusion for excitatory (A,B) and inhibitory (A*,B*) population. Data was taken from 2 simulations and $\mathrm{t=1000-1500\,s}$.}
\label{Fir_Rate_Dist_Instant_compare}
\end{figure}

Summing up the results of this section, we can state that stable network activity can be achieved in the LIF-SORN with diffusive homeostasis while producing a broad distribution of firing rates within the excitatory population. We found that this ensemble of activity is maintained on a single cell level, each cell spiking with approximately poissonian statistics at a constant mean rate. The broadness of distribution could be maintained at a relatively constant level over a wide range of diffusion rates. However, we did not find a distribution as right skewed and heavy tailed as reported by Sweeney et al. and numerous experimental studies.

\subsection{Network Topology}
So far, we have only analyzed network activity, whereas in this section we present results that relate to features of synaptic topology. Since the SORN we used for all simulations only included plastic mechanisms affecting recurrent excitatory connections, these were the main subject of our investigations. Moreover, in contrast to the previous section, we did not aim for the emergence of an explicitly \emph{new} feature of synaptic topology, but rather sought to recover properties that had been found in earlier versions of the LIF-SORN.

\subsubsection{Excitatory Connection Fraction}
In the previous version of the SORN, initializing the network with zero recurrent excitatory connections led to a monotonically increasing but saturating excitatory-to-excitatory connection fraction (CF). The growth rate was tuned such that the terminal CF settled at $\mathrm{10\%}$, see \cite[p. 8]{SORN_Paper}. Omitting the distance dependence of connection probability for synaptic growth led to a slightly higher connection fraction. To compare these previous results to the diffusive case, we ran a full simulation under diffusive homeostasis, including the growth phase by setting the target concentration to an appropriate value that was determined in a previous simulation with non-diffusive homeostasis. The resulting time course of CF is depicted in Figure \ref{CF_plot}.
%%% plot_CF.py
\includegraphics[width=\textwidth]{../../plots/CF_plot.png}
\caption{Connection fractions of recurrent excitatory synapses. Simulation protocols include all four possible combinations of diffusive/non-diffusive homeostasis and distance-dependent/non-distance-dependent synaptic growth, each variant retained for the entire simulation time.}
\label{CF_plot}
To our surprise, diffusive homeostasis caused the CF to overshoot its final fraction in the growth phase. Eventually however, it settled at a similar equilibrium of $\mathrm{\approx 10\%}$. To understand this kind of behaviour, we recalled the two mechanisms that directly determine changes within the connection fraction, namely synaptic growth and pruning. The average change of connection fraction is simply proportional to the difference between newly created synapses and those that were removed. Though fluctuating, the synaptic growth rate was kept at a constant level of 920 synapses/second. In consequence, any differences in Figure \ref{CF_plot} must originate in different pruning rates. As explained in section \ref{network simulation}, pruning is performed once per second by collection connections that have fallen under the given weight-threshold. Within this second, a synaptic weight can fall below this value for two reasons: either due to multiplicative normalization or additive STDP. If the weight of a neuron is already very small, multiplicative changes due to normalization have a minor effect compared to weight fluctuations caused by additive STDP, drawing us to the conclusion that differences within the statistics of STDP are the potential cause for the observed overshooting of the CF. Figure \ref{STDP_change} shows the mean and standard deviation of weight changes due to STDP between two normalization steps. Non-diffusive homeostasis has a higher mean as well as a broader distribution of weight fluctuations throughout the beginning of the simulation. However, the difference of mean changes is orders of magnitude smaller than the width of fluctuations. Thus, while it might seem counter-intuitive that a higher positive mean weight change comes with a smaller chance of synaptic survival, we attribute the main cause of overshooting in Figure \ref{CF_plot} to the increased fluctuation of weights, since it raises the chance of going below the pruning threshold within a given time interval.
%%% stdp_change_time.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/av_stdp_change_time.png}
\caption{Top: Mean change of synaptic weights (growth and pruning excluded) between two normalization steps. Bottom: Standard deviation for the same data. Both curves correspond to those in Figure \ref{CF_plot} with the same color.}
\label{STDP_change}
Apart from reproducing the desired connection fraction, previous versions of the LIF-SORN showed an over-representation of bidirectional connections compared to a random graph with equal connectivity \cite{SORN_Paper}. This experimentally observed feature \cite{Markram_Connections_1997,Song_Connectivity_2005} required the presence of a distance dependent connection probability, as described in section \ref{network simulation}. The absence of this breaking of spatial homogeneity even led to an under-representation of bidirectional connections, which is known to be an effect of STDP in recurrent networks \cite{Syn_Plast_Abbott}. As shown in Figure \ref{RC_CFB_plot}, the separation between simulations with and without a spatial connection profile is retained, but a slight shift towards higher fractions of bidirectional connections is preserved over a long time. 
%%% plot_RC_CFB.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/RC_CFB_plot.png}
\caption{Fraction of bidirectional connections normalized over the expected fraction for a random graph with equal total connection fraction.}
\label{RC_CFB_plot}
\subsubsection{Distribution of Synaptic Weights}
A very well-studied property of neural networks is the distribution of synaptic weights. Experimental studies in hippocampus and cortical regions suggest heavy-tailed log-normal-like distribution of synaptic efficacies \cite{Song_Connectivity_2005,Lisman_Synapses_1993,Yasumatsu_Synapses_2008,Loewenstein_Spine_Sizes}. Theoretical models are mainly based on a combination of multiplicative and additive weight dynamics \cite{Loewenstein_Spine_Sizes,Statman_Synapses_2014}, which is in line with our implementation of multiplicative normalization and additive STDP. Figure compares the resulting distributions. We found that we can retain a log-normal-like distribution in combination with diffusive homeostasis. A slightly larger variance can be observed. A possible explanation refers back to our observations in Figure \ref{STDP_change}: regarding STDP as a random process, a broader distribution of weights can be explained by a smaller potentiating effect and a decreased variance of STDP-fluctuations, which is both the case for diffusive homeostasis in the beginning of the simulation. Still, it seems unlikely that a possible effect of these differences is still present at $\mathrm{t=1500\,s}$.
%%% plot_weight_dist.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/syn_weight_dist.png}
\caption{Distribution of decadic logarithm of excitatory synaptic weights at $\mathrm{t=1500\,s}$ for diffusive and non-diffusive homeostasis (both with distance-dependent connectivity). Gaussian fits returned $\mathrm{\mu_{diff}=-0.172\, log_{10}(mV)}$, $\mathrm{\mu_{non-diff}=-0.087\, log_{10}(mV)}$, $\mathrm{\sigma_{diff}=0.643\, log_{10}(mV)}$ and $\mathrm{\sigma_{non-diff}=0.418\, log_{10}(mV)}$.}
\label{Weight_Dist}
\subsubsection{Synaptic Lifetimes}
Synaptic lifetimes have been shown to approximately follow a power law distribution in earlier versions of the LIF-SORN and binary SORN \cite{SORN_Paper,Pengsheng_2013}.
